---
title: "ETC5521 Tutorial 7"
subtitle: "Going beyond two variables, exploring high dimensions"
author: "Brendi Ang"
date: "Week 7"
output:
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  cache = TRUE,
  cache.path = "tutorial08-cache/",
  fig.path = "images/tutorial08/",
  fig.align = "center"
)

library(tidyverse)
library(ggExtra)
library(patchwork)
library(plotly)
library(GGally)
library(tourr)
library(RColorBrewer)
library(vcd)
library(cassowaryr)
library(spinifex)
library(vaast)

ggplot2::theme_set(theme_bw())
```

# 🖍 Objectives

These are exercises in plots to make to explore relationships between multiple variables. You will use interactive scatterplot matrices, interactive parallel coordinate plots and tours to explore the world beyond 2D.

# 🎨 Preparation 

Install the following R-packages if you do not have it already:

```{r pkgs, echo=TRUE, eval= FALSE}
install.packages(c("tidyverse", "ggExtra","patchwork", "GGally", "tourr", "spinifex", "plotly", "vcd", "SMPracticals", "vcdExtra", "RColorBrewer", "ggpubr"))

install.packages("cassowaryr")
```

# Exercise 1:  Melbourne housing

## a. 

Read in a copy of the Melbourne housing data from  [Nick Tierney's github repo](https://github.com/njtierney/melb-housing-data) which is a collation from the version at [kaggle](https://www.kaggle.com/anthonypino/melbourne-housing-market/version/21). Its fairly large, so let's start simply, and choose two suburbs to focus on. I recommend "South Yarra" and "Brighton". (Note: there are a number of missing values. I recommend removing these before making plots.)

```{r}
# South Yarra and Brighton
mel_houses <- read_csv("https://raw.githubusercontent.com/njtierney/melb-housing-data/master/data/housing.csv") %>%
  filter(suburb %in% c("South Yarra", "Brighton")) %>%
  # Remove missing values
  filter(across(.cols = c(bedroom2, bathroom, price), ~ !is.na(.x)))
```

## b.

Make a scatterplot matrix of `price`, `rooms`, `bedroom2`, `bathroom`, `suburb`, `type.` The plot will be easier to read if you put the numerical variables first, and then the categorical variables. What are the associations that can be seen?

```{r}
GGally::ggpairs(mel_houses, 
                columns = c(
                  "price", "rooms", "bedroom2", "bathroom", # numerical
                  "suburb", "type") # categorical
                ) 
```

* Positive linear association between price, rooms, bedroom2, bathroom, which indicates the bigger the house the higher the price
* From the boxplots: houses in Brighton tend to be higher priced and bigger than South Yarra, and houses tend to be worth more than apartments or units.
* From the fluctuation diagram, Brighton tends to have more houses, and South Yarra has more apartments.
* From the density plot, price has a skewed distribution.
* There is one big outlier, one property sold for a much higher price.

## c.

Subset the data to South Yarra only. Make an interactive scatterplot matrix of `rooms`, `bedroom2`, `bathroom` and `price`, coloured by type of property. There is a really high price property. Select this case, and determine what's special about it -- why did it sell for so much? Select the outlier in bedrooms and bathrooms, and examine the other characteristics of this property.

```{r}
south_yarra <- mel_houses %>% 
  filter(suburb == "South Yarra") %>% 
  select(rooms, bedroom2, bathroom, price, type)

plotly::highlight_key(south_yarra) %>%
  GGally::ggpairs(aes(color = type), 
                  columns = 1:4,
                  upper = list(continuous = "points")
                  ) %>% 
  ggplotly() %>%
  # Query when data points are clicked or selected
  highlight("plotly_selected")
```

* This property that has a high price has relatively modest characteristics! 
* The property with 5 bathrooms for 3 bedrooms is fairly low priced. Maybe there is a mistake in the data and the bedrooms/bathrooms were swapped.

**Code**

* `highlight_key()`: Controls how queries are triggered and rendered using the `highlight` function
* `highlight("plotly_selected")`: tells R that points are 'queried' when data point(s) are clicked or selected 

## d. 

Examine price vs rooms coloured by bathrooms, faceted by suburb and type, and with a linear model overlaid. What do you learn about average house prices relative to number of rooms and number of bathrooms, for the different property types and suburbs? (Remove the one really high priced property first, because it affects what we can learn about the rest of the data.)

```{r}
mel_houses %>%
  # remove outlier
  filter(price < 9000000) %>%
  ggplot(aes(x = rooms,
             y = price,
             colour = factor(bathroom))) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  colorspace::scale_colour_discrete_sequential(palette = "Viridis") +
  facet_grid(type ~ suburb)
```

* Three bathrooms, except for houses in Brighton seem to decrease prices as rooms increase!
* There are not many townhouses in South Yarra.
* For Brighton houses there is generally an increasing relationship between rooms and price as number of bathrooms increases.
* For both suburbs, generally units with 2 bathrooms are more highly priced, relative to rooms.

## e. 

If we throw all the neighbourhoods in together to analyse price and property characteristics, what paradox might we encounter?

```{r}
# SIMPSONS PARADOX
mel_houses %>% 
  ggplot() +
  geom_jitter(aes(x = rooms, y = price))
```

* Simpsons paradox -- statistical phenomenon in which overall trends or relationships between variables that exists either disappear or reverse when the data is broken down into groups or a new variable is introduced

* Simpsons paradox. Suburb is an important factor in property price. 
* The relationship between price and other characteristics are likely to be different by suburb, and this information will be lost.

# Exercise 2:  Olive oils

Following on from the olive oils example from class, we will explore the oils from the south here. 

## a.

Grab a copy of the data, and subset to contain just the samples from region = south (1), and also drop `eicosenoic` acid, because there is nothing useful about this variable for the southern oils.

```{r}
# South region 
olive <- read_csv("http://ggobi.org/book/data/olive.csv") %>%
  rename(id = `...1`) %>% 
  # South region
  filter(region == 1) %>%
  select(palmitic:arachidic, id, area) %>%
  mutate(area = as.integer(area))
```

## b.

Only looking at areas (1-3), that is not Sicily:

### i. 

Make an interactive parallel coordinate plot of the fatty acids (except `eicosenoic`), where the lines are coloured by area. 

```{r}
# Set our colours for the par coords, and also tour
# first 4 hex colours from the dark 2 palette
clrs <- brewer.pal(n = 4,
                   name = "Dark2") 

# Only three areas
not_sicily <- olive %>%
  filter(area != 4)

not_sicily_pcp <- 
  not_sicily %>%
  plot_ly(type = 'parcoords', 
          line = list(color = ~ area,
                      # Set own colour scale; e.g. 0 - 0.5 will be clrs[1]
                      colorscale = list(c(0, clrs[1]), 
                                        c(0.5, clrs[2]),
                                        c(1, clrs[3]))),
          # Add dimensions 
          dimensions = list(
            list(range = c(~ min(palmitoleic), ~ max(palmitoleic)),
                 label = 'palmitoleic', 
                 values = ~ palmitoleic),
            # negative values
            list(range = c(~ -max(oleic), ~ -min(oleic)),
                 label = 'oleic',
                 values = ~(-oleic)),
            list(range = c(~ min(palmitic), ~ max(palmitic)),
                 label = 'palmitic',
                 values = ~ palmitic),
            list(range = c(~ min(linoleic), ~ max(linoleic)),
                 label = 'linoleic',
                 values = ~ linoleic),
            # negative values
            list(range = c(~ -max(linolenic), ~ -min(linolenic)),
                 label = 'linolenic',
                 values = ~ (-linolenic)),
            # negative values
            list(range = c(~ -max(arachidic), ~ -min(arachidic)),
                 label = 'arachidic',
                 values = ~ (-arachidic)),
            # negative values
            list(range = c(~ - max(stearic), ~ - min(stearic)),
                 label = 'stearic',
                 values = ~ (-stearic))
            )
          ) 

not_sicily_pcp
```

    
**About the code**

  * Notice that some variables have been flipped by putting a minus sign in front, this is to make the correlation between variables positive, making the parallel coordinates easier to read.
  * Note the strange colour mapping: colour value has to range between 0-1 weirdly enough, so area 1 becomes 0, area 3 becomes 1. The range of each variable needs to be stated so that each is scaled from min to max for the display

**About the interactivity**

  * Click and drag along an axis to select observations
  * Click and drag the variable label to re-order. I find the best order to see groups is palmitoleic, oleic, palmitic, linoleic, linolenic, arachidic, stearic

```{r}
# All four areas
olive_pcp <- olive %>%
  plot_ly(type = 'parcoords', 
          line = list(color = ~area,
                      colorscale = list(c(0,clrs[1]), 
                                        c(0.33,clrs[2]),
                                        c(0.67,clrs[3]),
                                        c(1,clrs[4]))),
          dimensions = list(
            list(range = c(~ min(palmitoleic), ~ max(palmitoleic)),
                 label = 'palmitoleic',
                 values = ~palmitoleic),
            # negative values
            list(range = c(~ - max(oleic), ~ -min(oleic)),
                 label = 'oleic',
                 values = ~(-oleic)),
            list(range = c(~ min(palmitic), ~ max(palmitic)),
                 label = 'palmitic', 
                 values = ~palmitic),
            list(range = c(~ min(linoleic), ~ max(linoleic)),
                 label = 'linoleic', 
                 values = ~linoleic),
            # negative values
            list(range = c(~ -max(linolenic), ~ -min(linolenic)),
                 label = 'linolenic', 
                 values = ~(-linolenic)),
            # negative values
            list(range = c(~ - max(arachidic), ~ - min(arachidic)),
                 label = 'arachidic',
                 values = ~(-arachidic)),
            # negative values
            list(range = c(~ - max(stearic), ~ - min(stearic)),
                 label = 'stearic',
                 values = ~(-stearic))
              ))
olive_pcp
```


### ii.

Look at the data in a tour. Describe what you learn about differences between the three areas, whether these are separated. Are some variables more useful for distinguishing the three areas? Are there any outliers? 
    
```{r}
# Make a tour
# To do this live

set.seed(11111)

# Apply colours to each of the three areas
col <- clrs[not_sicily$area] 

# Grand tour
animate_xy(not_sicily[, 1:7],
           fps = 10,
           half_range = 1.0,
           col = col
           )

# Guided tour
animate_xy(not_sicily[,1:7],
           guided_tour(lda_pp(not_sicily$area),  
                       search_f = search_geodesic),
           fps = 10, 
           half_range = 1.0,
           col = col)

# All of areas
set.seed(11111)

# Apply colours to all areas
col <- clrs[olive$area]

# Grand tour on all areas
animate_xy(olive[, 1:7],
           fps = 10,
           half_range = 1.0,
           col = col)
```

```{r}
ggplot(data = not_sicily,
       aes(x = linoleic, y = oleic, colour = factor(area))) +
  geom_point()

ggplot(data = not_sicily,
       aes(x = palmitoleic, fill = factor(area))) +
  geom_density()
```

**Motivation for tour**
 
* Display low-dimensional projections from high-dimensional data, to explore the shape of multivariate distributions
  * identify outleirs when they're separated from other observations or move on diffeernt paths from other points
  
**Different tours**

* *Grand tour*: Projections (or bases) are randomly selected, and gives a global overview of the distribution
  * A "movie" of low-dimensional projections that comes arbitrarily close to showing all possible low-dimensional projections
  
* *Guided tour*: Projections is optimised using an index function
  * Tour moves towards more interesting views of the distribution as the animation progresses
  * "Interestingness" is measured by index function, that gets maximised along the tour path

4 indicies
1. LDA (`lda_pp`)
2. holes
3. central mass
4. PDA


## c. 

Re-do b. with Sicily. Explain what you learn about Sicily relative to the other areas.

* Do this at your own time!
* You'd notice that Sicily overlaps with two of the other three areas. Most of the samples are not distinguishable from the other two.

## d. 

Do some Googling. What can you find out about Sicilian olive oils? Are they higher in value? Does Sicily even grow olives, or does it use olives from neighbouring areas?

* Not sure what the reason is! 
* Maybe growing conditions in some fields are similar in the three areas. Maybe Sicily imports olives to make the oils from neighbouring areas.

# Exercise 3: Baker field soils

## a.

Make density plots of the soil variables in the Baker field corn yield data. Choose an appropriate transformation to symmetrise the distribution.

```{r fig.width=6, fig.height=4, eval=TRUE}
corn <- read_csv("https://eda.numbat.space/data/baker.csv")

# Long format for plotting
corn_long <- corn %>%
  select(B:Zn) %>%
  pivot_longer(everything(),
               names_to = "var",
               values_to = "value")

# Density plot
ggplot(corn_long, aes(x = value)) +
  geom_density() +
  facet_wrap( ~ var, ncol = 4, scales = "free")
```

* Many of the variables have a right-skewed distribution. 
* For this thinking about a square root or log transformation would be appropriate. Ca has a severe right-skew, so a log-log transformation might be required. The code below shows the transformations made. 

```{r echo=TRUE, eval=TRUE}
# Transformation
corn_trf <- corn %>%
  mutate(B = log10(B),
         Ca = log10(log10(Ca)), # log-log transformation for severe right-skew
         Cu = sqrt(Cu),
         K = log10(K), 
         Mg = sqrt(Mg), 
         Mn = sqrt(Mn),
         Na = sqrt(Na),
         P = log10(P),
         Zn = sqrt(Zn))

# Long format for plotting
corn_trf_long <- corn_trf %>% 
  select(B:Zn) %>%
  pivot_longer(everything(), 
               names_to="var",
               values_to="value")

# Density plot of transformed variable
ggplot(corn_trf_long, aes(x = value)) +
  geom_density() +
  facet_wrap( ~ var, ncol = 4, scales = "free")
```

## b.

Make a scatterplot matrix. If you can make an interactive one, that would be extra special. Describe the relationships between pairs of variables.

```{r fig.width=6, fig.height=6, eval=TRUE}
ggpairs(corn_trf, columns = 4:13)
```

## c.

Make a grand tour of soil variables. Describe the different patterns that you see in various projections. Is there clustering? Is there linear dependence? Non-linear dependence? outliers. For any structure that you see determine which variables contribute to it, and make plots of these variables (or check the scatterplot matrix) to check whether the pattern is visible there too.

```{r echo=TRUE}
set.seed(11111)

# Grand tour
animate_xy(corn_trf[, 4:13],
           fps = 10,
           half_range = 1)
```

```{r fig.width=6, fig.height=2, eval=TRUE}
# Check scatter plot matrix of interesting structure from the tour above

# Outliers - but this shows quite an odd relationship
p1 <- ggplot(corn_trf, aes(x = Ca, y = Mg)) +
  geom_point() + 
  theme(aspect.ratio = 1) 

# Nonlinear?
p2 <- ggplot(corn_trf, aes(x = Ca, y = Zn)) +
  geom_point() + 
  theme(aspect.ratio = 1) 

p3 <- ggplot(corn_trf, aes(x = Cu, y = Fe)) +
  geom_point() +
  theme(aspect.ratio = 1)  

p1 + p2 + p3 + plot_layout(ncol = 3)
```

# Exercise 4: Exam marks

There is a dataset "mathmarks" in the SMPracticals package, which has marks out of 100 for 88 students. It is interesting to note that all students had marks for all tests, which makes one wonder whether marks for students who missed a test were dropped. Mechanics and vectors were closed book exams, and the others were open book.

## a.

Make a side-by-side boxplot of the test scores. What do you learn  about  the test scores on the different subjects?

```{r}
data(mathmarks, package = "SMPracticals")

mathmarks_df <- mathmarks %>%
  # Denote if test is open or closed book
  rename_with(.cols = c("mechanics", "vectors"),
              .fn = ~ paste(.x, ("(C)"))) %>% 
  rename_with(.cols = c("algebra", "analysis", "statistics"),
              .fn = ~ paste(.x, ("(O)"))) 

mathmarks_long <- mathmarks_df %>% 
  pivot_longer(everything(), 
               names_to = "var",
               values_to = "marks") %>% 
  mutate(var = factor(var, levels = unique(var))) %>% 
  ggplot(aes(x = var, y = marks)) + 
  geom_boxplot()
```

There is some difference in median and IQR across exams. Students tended to do better on vectors, algebra and analysis and worse on mechanics and statistics. There is no indication that the open book exams produced better scores than the closed book exams.

The distributions are fairly symmetric. The spread is similar except for algebra which are more concentrated. There is one quite high score on algebra, and two low scores. Mechanics and vectors has an unusually low score each.

## b.

Make a scatterplot matrix, even better if it is interactive. Describe the relationships between the tests. Is there something different about the open book vs closed book scores?

```{r fig.width=6, fig.height=6, eval=TRUE}
ggpairs(mathmarks_df)
```

Generally we can see positive linear association. Algebra and analysis are most related. Algebra is pretty closely related to all other test scores. Mechanics is only weakly related to statistics. There are a couple of outliers: low scores on algebra, and also two students who did badly on vectors but quite well on analysis. There looks to be a barrier at 80 for the statistics test, maybe this was the maximum possible score. Analysis and algebra, and also analysis and vectors has what I call a "comet" distribution: there is a tight concentration of high scores, and a big spread among lower scores. I see this a lot in statistics test scores.

## c. 

Make an interactive parallel coordinate plot. Are there some students who have done consistently well on all tests? Consistently badly on all tests?  Badly on some but better on others?

```{r fig.width=6, fig.height=4, eval=TRUE}
marks_pcp <- mathmarks %>%
  plot_ly(type = 'parcoords', 
          dimensions = list(
            list(range = c(0,100),
                 label = 'Mechanics (C)', values = ~ mechanics), 
            list(range = c(0,100),
                 label = 'vectors (C)', values = ~ vectors), 
            list(range = c(0,100),
                 label = 'algebra (O)', values = ~ algebra), 
            list(range = c(1,100),
                 label = 'analysis (O)', values = ~ analysis), 
            list(range = c(1,100),
                 label = 'statistics (O)', values = ~ statistics)
            )
          )
marks_pcp
```

There is one student who did much worse on the closed book exams. There is one student who did really well on the closed book vectors exam, but badly on all other exams. There are a few students who have done really well on all exams. Generally, a student does relatively as well across all the tests. There appears to be some slight bimodality in mechanics, analysis and statistics which is also visible in the density plots in the splom.

# Exercise 5: Knowledge and resources

The "vcdExtra" package contains a dataset "Dyke" about how 1729 survey  respondents' knowledge of cancer depended on whether they listened to the radio, read newspapers, did solid reading, or attended lectures.

```{r}
data(Dyke, package = "vcdExtra")

# Convert array to tibble
Dyke_tsb <- as_tibble(Dyke) %>%
  mutate(Knowledge = factor(Knowledge, levels=c("Good", "Poor")),
         Reading = factor(Reading, levels=c("Yes", "No")),
         Radio = factor(Radio, levels=c("Yes", "No")),
         Lectures = factor(Lectures, levels=c("Yes", "No")),
         Newspaper = factor(Newspaper, levels=c("Yes", "No"))) 
```

## a.

Make separate bar charts for each of the explanatory variables, with bars filled by the response variable `Knowledge`. What do you learn?

```{r}
p1 <- ggplot(Dyke_tsb, aes(x = Reading, 
                           y = n,
                           fill = Knowledge)) + 
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_brewer(palette="Dark2")

p2 <- ggplot(Dyke_tsb, aes(x = Radio, 
                           y = n,
                           fill = Knowledge)) + 
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_brewer(palette="Dark2")

p3 <- ggplot(Dyke_tsb, aes(x = Lectures, 
                           y = n, 
                           fill = Knowledge)) + 
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_brewer(palette="Dark2")

p4 <- ggplot(Dyke_tsb, aes(x = Newspaper,
                           y = n,
                           fill = Knowledge)) + 
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_brewer(palette="Dark2")

p1 + p2 + p3 + p4 + plot_layout(nrow = 2)
```

## b.

Make a 100% bar chart of Newspaper, with Knowledge mapped to fill, and faceted by Reading. What do you learn about the relative proportions in the groups?

```{r fig.width=8, fig.height=4, eval=TRUE}
ggplot(Dyke_tsb, aes(x = Newspaper, 
                     y = n, 
                     fill = Knowledge)) +
  geom_bar(stat = "identity",
           position = position_fill(reverse = TRUE)) +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~ Reading, labeller = labeller(Reading = label_both))
```


## c.

Make a doubledecker plot of the data. What combination of factors leads to the highest level of knowledge about cancer?  What combination leads to the lowest?

```{r knowledge, eval = FALSE}
Dyke_tbl <- Dyke_tsb %>%
  mutate(Knowledge = factor(Knowledge, levels=c("Poor", "Good"))) %>%
  uncount(n) %>%
  select(Reading, Radio, Lectures, Newspaper, Knowledge) %>%
  arrange(Knowledge, Reading, Radio, Lectures, Newspaper)

doubledecker(Knowledge ~ .,
             # Radio + Lectures,
             data = Dyke_tbl,
             gp = gpar(fill = c("grey90", "orangered")))
```

The most knowledgeable people read, listen to the radio, attend lectures and read newspapers - as might be expected. There a very few people in this category, though. On each comparison the Yes group is more knowledgeable than the No group. The most populous group is the No to all combination. This is pretty shocking because it is the least knowledgeable about cancer. The second most common group are people who read newspapers, and read generally, and these are relatively knowledgeable about cancer.

# Exercise 6: Parkinsons

This dataset is composed of a range of biomedical voice measurements from 31 people, 23 with Parkinson's disease (PD). Each column in the table is a particular voice measure, and each row corresponds one of 195 voice recording from these individuals ("name" column). The main aim of the data is to discriminate healthy people from those with PD, according to "status" column which is set to 0 for healthy and 1 for PD. 

The data is available at [The UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Parkinsons) in ASCII CSV format. The rows of the CSV file contain an instance corresponding to one voice recording. There are around six recordings per patient, the name of the patient is identified in the first column. There are 24 variables in the file, including the persons name in column 1. 

The data are originally analysed in:
Max A. Little, Patrick E. McSharry, Eric J. Hunter, Lorraine O. Ramig (2008), 'Suitability of dysphonia measurements for telemonitoring of Parkinson's disease', IEEE Transactions on Biomedical Engineering (to appear).

## a.

Compute several of the scagnostics (monotonic, outlying) for all the first 5 variables, except for `name` and `status`. 

```{r eval=TRUE}
# Read the data
data(pk, package = "cassowaryr")

# Compute scagnostics
s <- calc_scags_wide(pk[, 2:5], scags = c("outlying", "monotonic", "sparse"))
```

## b. 

Sort the scagnostics, by (i) Monotonic (ii) Outlying (iii) Your choice, and plot the pair of variables with the highest values. 

```{r}
s %>% 
  arrange(desc(sparse)) 

# Highest monotonic values
ggplot(data = pk, 
       aes(x = `MDVP:Fhi(Hz)`, y = `MDVP:Fo(Hz)`)) + 
  geom_point() 

# Highest outlying values
ggplot(data = pk,
       aes(x = `MDVP:Jitter(%)`, y = `MDVP:Fhi(Hz)`)) +
  geom_point()

# Highest sparse values
ggplot(data = pk,
       aes(x = `MDVP:Jitter(%)`, y = `MDVP:Fo(Hz)`)) +
  geom_point()
```

* There are some variables that are exactly the same, up to a scale factor.


## c. 

Make an interactive scatterplot matrix. Browse over it to choose other interesting pairs of variables and make the plots.

```{r}
# Create an interactive splom
s <- s %>%
  mutate(vars = paste(Var1, Var2))

highlight_key(s) %>%
  GGally::ggpairs(columns = 3:5, 
                  mapping = aes(label = vars)) %>%
  ggplotly(tooltip = "all") %>%
  highlight("plotly_selected")
```

```{r eval=TRUE, fig.width=3, fig.height=3}
# Plot some of the pairs that have high scagnostics values
ggplot(data = pk, aes(x = `MDVP:PPQ`, y = `Jitter:DDP`)) +
  geom_point() # High on monotonic and outlying
```

* This pair of variables has both positive association, and outliers.


## d.

The scagnostics help us to find interesting associations between pairs of variables. However, the problem here is to detect differences between Parkinsons patients and normal patients. How would you go about that? Think about some ideas long the line of scagnostics but look for differences between the two groups.

```{r}
# One way to examine difference between Parkinsons and healthy
pk_med <- pk %>% 
  select(-name) %>% 
  group_by(status) %>%
  summarise_all(list(median, sd)) %>%
  pivot_longer(cols = `MDVP:Fo(Hz)_fn1`:`PPE_fn2`,
               names_to = "var",
               values_to = "value") %>% 
  separate(var, c("var","stat"), "_") %>%
  mutate(stat = fct_recode(stat,
                           "m" = "fn1",
                           "s" = "fn2")) %>%
  pivot_wider(names_from = stat,
              values_from = value) %>% 
  group_by(var) %>%
  summarise(d = (m[status == 0] - m[status == 1]) / sqrt(s[status == 0] ^ 2 +
                                                           s[status == 1] ^ 2))

ggplot(pk, aes(x = factor(status), y = `MDVP:Fo(Hz)`)) + 
  geom_boxplot()
```

Generally we are looking for variables where the differences between the Parkinsons and normal patients are big. You need to measure big, relative to the variance of each group. Doing a two sample t-test for each variable is one approach. Here, I've computed the median for each group of patients and compared the difference in medians relative to the pooled standard deviation in each group.
